Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-07-27 paper-amia-summit-2018/manuscript.tex
update data collection and Features sections

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced lexical, punctuation and  part-of-speach (POS) features and experimented with both traditional supervised machine learning method, linear-chain conditional random fields (CRF) and deep learning methods, such as multilayer perceptrons (MLP), bidirectional recurrent neural networks (BRNN) and convolutional recurrent neural networks (CRNN), to find the best performing method and feature combination.

Relevant previous work in the biomedical domain primarily focused on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence,fraser2015sentence}. Apostolova et al.\cite{apostolova2009automatic} applied SVM along with word-vector cosine similarity metric combined with several heuristics to segment clinical reports into sections, such as demographics, history, procedure, finding and impression. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for section classification. Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules and Naive Bayes classifier to identify the sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM based on prosodic and part of speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks using word embeddings \cite{griffis2016quantitative} were utilized for detecting sentence boundaries. Liu et al. have shown that a linear chain CRF performed better than hidden markov model and Maxent on the NIST sentence boundary detection \cite{liu2005using}. Recently, Treviso et al. proposed a recurrent convolutional neural networks for automatic identification of sentence boundary in neuropsychological text in Portuguese language \cite{treviso2017sentence}. Their proposed model combined prosodic, part of speech and lexical features, which achieved 0.74 F1-score in sentence boundary detection in Brazilian constitution dataset. We adapted this model in our work, which was employed differently from the aforementioned studies. Our study focus on segmentation, rather than determining the sentence boudary. Another difference is that punctuation features were utilized in combination with part-of-speech features. Besides, our model trains using 300 dimension pre-trained word2Vec word vectors \cite{mikolov2013distributed}, rather than utilizing 50 dimension pre-trained GloVe word vectors \cite{pennington2014glove}. Our experiments show that these differences lead to significant improvement in the segmenteation of e-Coaching emails. Segmentation of e-Coaching emails is also different from a traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers.

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We formulate the segmentation task as a binary classification problem where each word annotated with a label 1 or 0, to indicate whether it preceds a new segment or not. We experimented both with traditional machine learning method, linear-chain Conditional Random Field (CRF) and deep learning methods, such as Multilayer Perceptrons (MLPs), Bidirectional Recurrent Neural Networks (BRNNs) and Convolutional Recurrent Neural Networks (CRNNs). For MLP models, e-Coaching email exchanges are partitioned into adjacent word pairs. Each pair is classified into either ``new segment'' or ``same segment'' class. The gold standard is manually segmented where adjacent word pairs are classified into ``new segment'' class. If all adjacent word pairs of a block of text are classified into ``same segment'' class, the entire block is treated as one textual fragment corresponding to a single MI behavior. In total, we obtained 95,421 word pairs, which include 3,138 ``new segment'' and 92,283 ``same segment'' instances. For deep learning models, an email was taken as input sequence, such that PoS tags and word embedding of each word or punctuation marks were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of the model at each step. In the gold standard, words within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: lexical, punctuation and topic features. Lexical features represent each word in a pair as a binary vector, in which 1 corresponds to the word in question and 0 to all other words. Since part-of-speech (PoS) tags \cite{kotov2015interpretable,hashimoto2016topic,lu2016modeling} have been shown to be effective semantic abstractions of individual words, we used topic features for our experiment. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to segment boundary.

@@

We experimented both with traditional classifiers, including Conditional Random Field (CRF) and deep learning models, such as Multilayer Perceptrons (MLPs) Bidirectional Recurrent Neural Networks (BRNNs) and Convolutional Recurrent Neural Networks (CRNNs).

@@

\textbf{Deep Learning Methods}: this method considers each training sample as a point in the input feature space. For a new test sample, Euclidean distance is calculated to find the k-nearest classified neighbors from the training set. Finally, the test sample is classified into a majority class of its k-nearest neighbors. We experimentally determined that best performance was achieved when the number of neighbors is 3.

@@

>>>>

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced lexical, punctuation and  part-of-speech (POS) features and experimented with both traditional supervised machine learning method, linear-chain conditional random fields (CRF) and deep learning methods, such as multilayer perceptrons (MLP), bidirectional recurrent neural networks (BRNN) and convolutional recurrent neural networks (CRNN), to find the best performing method and feature combination.

Relevant previous work in the biomedical domain primarily focused on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence,fraser2015sentence}. Apostolova et al.\cite{apostolova2009automatic} applied SVM along with word-vector cosine similarity metric combined with several heuristics to segment clinical reports into sections, such as demographics, history, procedure, finding and impression. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for section classification. Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules and Naive Bayes classifier to identify the sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM based on prosodic and part of speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks using word embeddings \cite{griffis2016quantitative} were utilized for detecting sentence boundaries. Liu et al. have shown that a linear chain CRF performed better than the hidden markov model and Maxent on the NIST sentence boundary detection \cite{liu2005using}. Recently, Treviso et al. proposed a recurrent convolutional neural network for automatic identification of sentence boundary in a neuropsychological text in Portuguese language \cite{treviso2017sentence}. Their proposed model combined prosodic and part of speech features with pre-trained GloVe word vectors \cite{pennington2014glove}, which achieved 0.74 F1-score in sentence boundary detection in the Brazilian constitution dataset.

In this study, we adopt the model proposed by Treviso et al. \cite{treviso2017sentence} and apply it differently from the aforementioned studies. The main difference is that our study focuses on segmentation of e-Coaching emails as distinct from determining the sentence boundary. Another important difference is that punctuation features were utilized in combination with word embeddings and part-of-speech features. Furthermore, we train our model with pre-trained word2vec word vectors \cite{mikolov2013distributed}, instead of using GloVe word vectors. Segmentation of e-Coaching emails is also different from a traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers.

%Our experiments show that these differences lead to significant improvement in the segmentation of e-Coaching emails.

@@

\section*{Methods}
\subsection*{\textit{Data collection}}

The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We formulate the segmentation task as a binary classification problem where each word or punctuation marks annotated with a label 1 or 0, to indicate whether it precedes a new segment or not. In total, we obtained 95,777 words and 7,140 punctuation marks, which include 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning method, CRF and deep learning methods, such as MLP, BRNN and CRNN. For MLP models, samples were created with $2l$ words/punctuation marks at each step in a word sequence, where a sample contains next $l$ words/punctuation marks and prior $l$ words/punctuation marks including current word/punctuation mark. Each sample is classified into either ``new segment'' or ``same segment'' based on whether the current word/punctuation mark precedes a new segment or not. For CRF, BRNN and CRNN models, an email was taken as input sequence, such that POS tags and word embedding of each word or punctuation marks were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of the model at each step. In the gold standard, words within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used topic features for our experiment \cite{liu2005using,treviso2017sentence}. To extract our POS features, we tag the e-Coaching emails using the NLTK POS tagger \cite{bird2009natural}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary. For natural language processing (NLP) tasks, inputs are received as a text where words can be considered as the basic processing unit. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embeddings are one of this representation where each word represented as a real-valued vector in a high dimensional vector space. The distributed representation of word allows capturing semantic, syntactic and morphological information from large unannotated corpora  \cite{pennington2014glove, mikolov2013distributed}. In our experiment, we utilized pre-trained word2vec word vectors for words and trained word2vec word vectors for punctuation marks.

@@

In the present study, we experimented with four different classifiers, including one traditional machine learning method, CRF and three neural networks models. Since neural networks provide a flexible architecture for constructing and synthesizing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

@@

\textbf{Multilayer Perceptrons}: this method considers each training sample as a point in the input feature space. For a new test sample, Euclidean distance is calculated to find the k-nearest classified neighbors from the training set. Finally, the test sample is classified into a majority class of its k-nearest neighbors. We experimentally determined that best performance was achieved when the number of neighbors is 3.

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods with different topic model-based features for segmentation of e-Coaching emails as a weighted average over ``same segment'' and ``new segment'' classification results, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\multirow{2}{*}{\textbf{Features}} & \multicolumn{4}{|c|}{\textbf{Area Under the Curve (AUC)}} \\\cline{2-5}
 & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings & \textbf{0.966} & 0.951 & 0.920 & 0.965 \\ \hline
 all features & \textbf{0.994} & 0.992 & 0.981 & 0.986 \\ \hline
  \end{tabular}
\end{table}
